---
title: "ANSWER KEY — Smoking-Attributable IHD Deaths in England"
subtitle: "INSTRUCTOR USE ONLY"
date: "2025"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    highlight: tango
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Data

```{r data}
smoking <- data.frame(
  year       = 2012:2023,
  prev_men   = c(21.9, 19.2, 18.4, 17.1, 15.9, 19.1, 14.7, 19.7, 16.3, 15.9, 17.7, 15.4),
  prev_women = c(17.4, 18.0, 17.3, 15.2, 13.2, 14.9, 15.9, 14.0, 15.6, 13.3, 12.2, 12.6)
)
smoking$p_men   <- smoking$prev_men   / 100
smoking$p_women <- smoking$prev_women / 100

ihd <- data.frame(
  year         = 2012:2023,
  deaths_men   = c(2884, 2914, 2855, 2915, 2948, 3033, 3131, 3173, 3477, 3568, 3632, 3882),
  deaths_women = c( 683,  713,  694,  724,  731,  740,  761,  774,  814,  853,  900,  959)
)

RR_men   <- 3.04
RR_women <- 3.78
```

---

# Question 1 — Completed code and answers

```{r q1-pif}
p_star <- c(0, 1)

results <- data.frame(
  year = 2012:2023,
  PIF_men = NA_real_, PIF_women = NA_real_,
  attr_men = NA_real_, attr_women = NA_real_
)

for (i in 1:nrow(results)) {
  p_men_obs   <- c(smoking$p_men[i],   1 - smoking$p_men[i])
  p_women_obs <- c(smoking$p_women[i], 1 - smoking$p_women[i])
  RR_men_vec   <- c(RR_men,   1)
  RR_women_vec <- c(RR_women, 1)

  results$PIF_men[i]   <- (sum(p_men_obs   * RR_men_vec)   - sum(p_star * RR_men_vec))   / sum(p_men_obs   * RR_men_vec)
  results$PIF_women[i] <- (sum(p_women_obs * RR_women_vec) - sum(p_star * RR_women_vec)) / sum(p_women_obs * RR_women_vec)

  results$attr_men[i]   <- results$PIF_men[i]   * ihd$deaths_men[i]
  results$attr_women[i] <- results$PIF_women[i] * ihd$deaths_women[i]
}

results$PIF_men    <- round(results$PIF_men   * 100, 1)
results$PIF_women  <- round(results$PIF_women * 100, 1)
results$attr_men   <- round(results$attr_men)
results$attr_women <- round(results$attr_women)
results$attr_total <- results$attr_men + results$attr_women

results
```

**Expected output (key rows):**

| Year | PIF men (%) | PIF women (%) | Attr men | Attr women | Total |
|---|---|---|---|---|---|
| 2012 | 30.7 | 34.2 | 886 | 234 | 1,120 |
| 2023 | 23.6 | 27.2 | 916 | 261 | 1,177 |

```{r q1-plot, fig.width=8, fig.height=5}
plot(results$year, results$attr_men,
     type = "b", pch = 16, lwd = 2, col = "#1E3A5F",
     ylim = c(0, 1400), xlab = "Year",
     ylab = "Smoking-attributable IHD deaths",
     main = "Smoking-attributable IHD deaths, ages 50–59, England")
lines(results$year, results$attr_women,
      type = "b", pch = 17, lwd = 2, lty = 2, col = "#2A9D8F")
lines(results$year, results$attr_total,
      type = "b", pch = 15, lwd = 2, lty = 3, col = "#E76F51")
legend("topleft", legend = c("Men", "Women", "Total"),
       col = c("#1E3A5F", "#2A9D8F", "#E76F51"),
       lwd = 2, lty = c(1,2,3), pch = c(16,17,15), bty = "n")
```

```{r q1-cumulative}
cat("Cumulative attributable IHD deaths, 2012–2023:\n")
cat("  Men:   ", sum(results$attr_men),   "\n")
cat("  Women: ", sum(results$attr_women), "\n")
cat("  Total: ", sum(results$attr_total), "\n")
```

**Expected cumulative totals:** Men ≈ 10,700; Women ≈ 2,870; Total ≈ 13,570.

---

**Q1.1 — Answer:**
Attributable deaths = PIF × total IHD deaths. The PIF is declining (because smoking prevalence is falling), but total IHD deaths rose markedly over the same period — from 3,567 in 2012 to 4,841 in 2023, a 36% increase. This rise is driven by an ageing population, rising obesity and diabetes, and a possible COVID-19-related increase in cardiovascular events from 2020 onwards. The rising death count more than offsets the falling PIF, so the product — attributable deaths — increases. This is an important public health lesson: reducing a risk factor does not automatically reduce attributable burden if the underlying disease is increasing.

**Q1.2 — Answer:**
Attributable deaths depend on both the PIF and the total number of deaths. Women have a higher PIF (their higher RR means each female smoker contributes more excess risk), but they have far fewer total IHD deaths than men — roughly 4:1 male:female ratio across all years. Since attributable deaths = PIF × total deaths, women's much smaller pool of IHD deaths produces fewer attributable deaths even with the higher PIF.

**Q1.3 — Answer:**
The cumulative total (~13,570 deaths) represents the estimated number of IHD deaths in 50–59 year olds in England over 12 years that would not have occurred if nobody had smoked — i.e., the total preventable burden attributable to current smoking in this age group. The key assumptions are: (1) the RR is constant over time and equally applicable in all years; (2) the ONS smoking prevalence accurately represents the true proportion of current smokers; (3) the GBD IHD death estimates are reliable; and (4) the causal effect assumption holds — that eliminating smoking would reduce IHD risk in proportion to the RR.

---

# Question 2 — Completed code and answers

```{r q2-scenario}
p_men_2023   <- smoking$p_men[smoking$year   == 2023]
p_women_2023 <- smoking$p_women[smoking$year == 2023]

p_men_cf   <- smoking$p_men[smoking$year   == 2012] / 2   # 0.1095
p_women_cf <- smoking$p_women[smoking$year == 2012] / 2   # 0.0870

deaths_men_2023   <- ihd$deaths_men[ihd$year   == 2023]
deaths_women_2023 <- ihd$deaths_women[ihd$year == 2023]

p_obs_men   <- c(p_men_2023,   1 - p_men_2023)
p_obs_women <- c(p_women_2023, 1 - p_women_2023)
p_cf_men    <- c(p_men_cf,     1 - p_men_cf)
p_cf_women  <- c(p_women_cf,   1 - p_women_cf)

RR_men_vec   <- c(RR_men,   1)
RR_women_vec <- c(RR_women, 1)

PIF_men_scenario   <- (sum(p_obs_men   * RR_men_vec)   - sum(p_cf_men   * RR_men_vec))   / sum(p_obs_men   * RR_men_vec)
PIF_women_scenario <- (sum(p_obs_women * RR_women_vec) - sum(p_cf_women * RR_women_vec)) / sum(p_obs_women * RR_women_vec)

attr_men_scenario   <- round(PIF_men_scenario   * deaths_men_2023)
attr_women_scenario <- round(PIF_women_scenario * deaths_women_2023)

attr_men_2023   <- results$attr_men[results$year   == 2023]
attr_women_2023 <- results$attr_women[results$year == 2023]

averted_men   <- attr_men_2023   - attr_men_scenario
averted_women <- attr_women_2023 - attr_women_scenario

cat("Attributable deaths (observed)  — Men:", attr_men_2023,     " Women:", attr_women_2023,     "\n")
cat("Attributable deaths (scenario)  — Men:", attr_men_scenario, " Women:", attr_women_scenario, "\n")
cat("Deaths averted                  — Men:", averted_men,       " Women:", averted_women,       "\n")
cat("Total averted:", averted_men + averted_women, "\n")
```

**Expected output:**
- Observed 2023: Men ≈ 916, Women ≈ 261, Total ≈ 1,177
- Scenario 2023: Men ≈ 472, Women ≈ 134, Total ≈ 606
- Averted: Men ≈ 444, Women ≈ 127, **Total ≈ 571**

```{r q2-plot, fig.width=7, fig.height=5}
heights <- c(attr_men_2023, attr_men_scenario, attr_women_2023, attr_women_scenario)
cols    <- c("#1E3A5F", "#7BA7C7", "#2A9D8F", "#8FD4CA")
labels  <- c("Men\n(observed)", "Men\n(scenario)", "Women\n(observed)", "Women\n(scenario)")

bp <- barplot(heights, names.arg = labels, col = cols, border = NA,
              ylab = "Smoking-attributable IHD deaths",
              main = "Observed vs. accelerated-decline scenario, 2023\nEngland, ages 50–59",
              ylim = c(0, max(heights) * 1.25))
text(x = bp, y = heights + 15, labels = heights, cex = 0.9)
```

---

**Q2.1 — Answer:**
Approximately 571 deaths would have been averted. Total IHD deaths in the 50–59 age group in 2023 were 4,841 (3,882 + 959). So 571 / 4,841 ≈ **12%** of all IHD deaths in this age group would have been prevented by the accelerated decline scenario. This is a substantial fraction attributable to the gap between actual and scenario prevalence in a single year in a single age group.

**Q2.2 — Answer:**
This is an overestimate of the averted deaths. The scenario assigns former smokers (those who quit under the accelerated decline) an RR of 1.0 — the same as never-smokers. In reality, former smokers retain elevated IHD risk for years after quitting; the risk declines gradually over roughly 10–15 years. Since their RR is above 1.0, the true reduction in the PIF from moving people from current to recent ex-smoker would be smaller than the formula assumes. The estimate is therefore biased upward — it overstates the deaths averted in the short run.

**Q2.3 — Answer (any one):**
- Total IHD deaths changed between 2012 and 2023 for reasons unrelated to smoking (obesity, diabetes, COVID-19 sequelae, population ageing). The analysis uses observed deaths in each year, so trends in background IHD risk are partly accounted for, but counterfactual IHD death counts under the scenario are not modelled.
- The RR may itself have changed over time as cigarette composition, smoking intensity, and co-exposures evolved.
- The analysis does not account for second-hand smoke exposure, which also contributes to IHD deaths in non-smokers.
- Survey non-response and self-report bias in smoking prevalence may have changed over time.

---

# References

1. Office for National Statistics. *Adult Smoking Habits in England, 2024.* Newport: ONS; 2025.
2. GBD 2023 Causes of Death Collaborators. *Global Burden of Disease Study 2023.* Seattle: IHME; 2024.
3. Ezzati M, Lopez AD, Rodgers A, Vander Hoorn S, Murray CJL. Selected major risk factors and global and regional burden of disease. *Lancet.* 2002;360(9343):1347–1360.
